quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-04-08 paper-quollr.Rmd
Find the best fit for the application dataset

```{r}
clust_df <- clusters_different_shapes_diff_num_points(sample_size = 1500, with_seed = NULL, cluster_size_vec = c(250, 150, 150, 150, 350, 450), num_gussian_clusters = 4, num_non_gaussian_clusters = 2, cluster_sd_gau = 0.2, cluster_sd_non_gau = 0.3, num_dims = 7, a = 2, b = 4)

@@

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
tsne_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tSNE_df, 
                                    x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tSNE_df))    
## tSNE
hex_size_vec <- seq(0.02, 2, by = 0.01)
vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names
mse_df_gau <- dplyr::bind_rows(vec)[0, ]
mse_df_gau <- mse_df_gau |>
  dplyr::mutate_if(is.character, as.numeric)
for (i in 1:length(hex_size_vec)) {

@@

}
### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_gau |>
  dplyr::filter(!(num_bins %in% dupli_bins))
### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_clust <- dplyr::bind_rows(not_dupli_df, duplicate_df) 
```

```{r}
mse_clust_plot <- ggplot(mse_df_clust, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("log(MSE)") +
  xlab("total number of bins")
```

```{r, fig.cap="Parameter tuning plots, where the best choice of parameter at a large drop in MSE. Here, these are at *total number of bins* = 180."}
mse_clust_plot
```

@@

To perform the algorithm on the clustering data, we first scale the tSNE data. We then let *number of bins along the x-axis* be $21$, *number of bins along the y-axis* be $28$, and *hex_size* be 0.03; see [Selecting parameter values] for further discussion and guidance regarding selection of these choices.

@@

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 21
num_bins_y_tsne_gau <- 28
hex_size_tsne_gau <- 0.03
## non-empty:198
hb_obj_tsne_gau <- hex_binning(data = tsne_gau_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "tSNE")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()
tsne_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$data_hb_id))

model_object_tsne_gau <- fit_highd_model( training_data = clust_df, 
                                 nldr_df_with_id = tsne_gau_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "x")
df_bin_centroids_tsne_gau <- model_object_tsne_gau$df_bin_centroids
df_bin_tsne_gau <- model_object_tsne_gau$df_bin
## Triangulate bin centroids
tr1_object_tsne_gau <- tri_bin_centroids(df_bin_centroids_tsne_gau, x = "c_x", y = "c_y")
tr_from_to_df_tsne_gau <- gen_edges(tri_object = tr1_object_tsne_gau)
## Compute 2D distances
distance_tsne_gau <- cal_2d_dist(tr_coord_df = tr_from_to_df_tsne_gau, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))
# distance_plot <- plot_dist(distance_tsne_gau) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot
## To find the benchmark value
benchmark_tsne_gau <- find_lg_benchmark(distance_edges = distance_tsne_gau, distance_col = "distance")
# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))
trimesh_removed_tsne_gau <- vis_rmlg_mesh(distance_edges = distance_tsne_gau, benchmark_value =
benchmark_tsne_gau, tr_coord_df = tr_from_to_df_tsne_gau, distance_col = "distance")
trimesh_removed_tsne_gau <- trimesh_removed_tsne_gau +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

df_b <- df_bin_tsne_gau |>
  dplyr::filter(hb_id %in% df_bin_centroids_tsne_gau$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_tsne_gau$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_tsne_gau |>
  dplyr::filter(distance < benchmark_tsne_gau)
## Since erase brushing is considerd.
langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(0.7, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))

<!-- model with bin means-->
```{r}
## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 21
num_bins_y_tsne_gau <- 28
hex_size_tsne_gau <- 0.03
## non-empty:198
hb_obj_tsne_gau <- hex_binning(data = tsne_gau_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start
= "tSNE")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" =
"hb_id"))
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()
tsne_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$data_hb_id))

model_object_tsne_gau <- fit_highd_model( training_data = clust_df, 
                                 nldr_df_with_id = tsne_gau_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_bin_centroid = FALSE,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "x")
df_bin_centroids_tsne_gau <- model_object_tsne_gau$df_bin_centroids
df_bin_tsne_gau <- model_object_tsne_gau$df_bin
## Triangulate bin centroids
tr1_object_tsne_gau <- tri_bin_centroids(df_bin_centroids_tsne_gau, x = "c_x", y = "c_y")
tr_from_to_df_tsne_gau <- gen_edges(tri_object = tr1_object_tsne_gau)
## Compute 2D distances
distance_tsne_gau <- cal_2d_dist(tr_coord_df = tr_from_to_df_tsne_gau, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))
# distance_plot <- plot_dist(distance_tsne_gau) +
  # ylab(expression(d^{(2)})) +
  # theme(axis.text = element_text(size = 5),
        # axis.title = element_text(size = 12))

# distance_plot
## To find the benchmark value
benchmark_tsne_gau <- find_lg_benchmark(distance_edges = distance_tsne_gau, distance_col =
"distance")

benchmark_tsne_gau <- 0.17

# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))
trimesh_removed_tsne_gau <- vis_rmlg_mesh(distance_edges = distance_tsne_gau, 
                                          benchmark_value = benchmark_tsne_gau, 
                                          tr_coord_df = tr_from_to_df_tsne_gau, distance_col = "distance")
trimesh_removed_tsne_gau <- trimesh_removed_tsne_gau +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size =
3)
trimesh_removed_tsne_gau
df_all_tsne_gau <- dplyr::bind_cols(clust_df |> dplyr::select(-ID),
tsne_data_with_hb_id_gau)
### Define type column
df <- df_all_tsne_gau |>
  dplyr::select(tidyselect::starts_with("x")) |>
  dplyr::mutate(type = "data") ## original dataset
df_b <- df_bin_tsne_gau |>
  dplyr::filter(hb_id %in% df_bin_centroids_tsne_gau$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_tsne_gau$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_tsne_gau |>
  dplyr::filter(distance < benchmark_tsne_gau)
## Since erase brushing is considerd.
langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)),
rep(0.7, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```

>>>>

```{r}
clust_df <- clusters_different_shapes_diff_num_points(sample_size = 1500, with_seed = NULL, cluster_size_vec = c(250, 150, 150, 150, 350, 450), num_gussian_clusters = 4, num_non_gaussian_clusters = 2, cluster_sd_gau = 0.2, cluster_sd_non_gau = 0.3, num_dims = 5, a = 2, b = 4)

@@

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
tsne_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tSNE_df, 
                                    x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tSNE_df))    
## Model with bin centroid
hex_size_vec <- seq(0.03, 2, by = 0.01)
vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names
mse_df_gau <- dplyr::bind_rows(vec)[0, ]
mse_df_gau <- mse_df_gau |>
  dplyr::mutate_if(is.character, as.numeric)
for (i in 1:length(hex_size_vec)) {

@@

}
### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_gau |>
  dplyr::filter(!(num_bins %in% dupli_bins))
### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_clust <- dplyr::bind_rows(not_dupli_df, duplicate_df) |>
  dplyr::mutate(type = "bin centroids")
```

mse_clust_plot <- ggplot(mse_df_clust, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 238, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("log(MSE)") +
  xlab("total number of bins")
```

```{r, fig.cap="Parameter tuning plots, where the best choice of parameter at a large drop in MSE. Here, these are at *total number of bins* = 238."}
mse_clust_plot
```

@@

To perform the algorithm on the clustering data, we first scale the tSNE data. We then let *number of bins along the x-axis* be $14$, *number of bins along the y-axis* be $27$, and *hex_size* be $0.05$; see [Selecting parameter values] for further discussion and guidance regarding selection of these choices.

@@

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 14
num_bins_y_tsne_gau <- 27
hex_size_tsne_gau <- 0.05
x_start_tsne_gau <- -0.0481
y_start_tsne_gau <- 0.00833
## non-empty:80
hb_obj_tsne_gau <- hex_binning(data = tsne_gau_scaled, x = "tSNE1", 
                      y = "tSNE2", num_bins_x = num_bins_x_tsne_gau, 
                      num_bins_y = num_bins_y_tsne_gau, x_start = x_start_tsne_gau, y_start = y_start_tsne_gau,
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_tsne_gau, col_start = "tSNE")
## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$centroids))
## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$hex_poly))
## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()
hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))
# ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
#   geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
#   geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
#   scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
#   coord_fixed()
tsne_data_with_hb_id_gau <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$data_hb_id))

model_object_tsne_gau <- fit_highd_model( training_data = clust_df, 
                                 nldr_df_with_id = tsne_gau_scaled, 
                                 x = "tSNE1", y = "tSNE2", 
                                 num_bins_x = num_bins_x_tsne_gau, 
                                 num_bins_y = num_bins_y_tsne_gau, 
                                 x_start = x_start_tsne_gau, y_start = y_start_tsne_gau, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_tsne_gau,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "tSNE", 
                                 col_start_highd = "x")
df_bin_centroids_tsne_gau <- model_object_tsne_gau$df_bin_centroids
df_bin_tsne_gau <- model_object_tsne_gau$df_bin
## Triangulate bin centroids
tr1_object_tsne_gau <- tri_bin_centroids(df_bin_centroids_tsne_gau, x = "c_x", y = "c_y")
tr_from_to_df_tsne_gau <- gen_edges(tri_object = tr1_object_tsne_gau)
## Compute 2D distances
distance_tsne_gau <- cal_2d_dist(tr_coord_df = tr_from_to_df_tsne_gau, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))
# distance_plot <- plot_dist(distance_tsne_gau) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot
## To find the benchmark value
benchmark_tsne_gau <- find_lg_benchmark(distance_edges = distance_tsne_gau, distance_col = "distance")

# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))
trimesh_removed_tsne_gau <- vis_rmlg_mesh(distance_edges = distance_tsne_gau, benchmark_value =
benchmark_tsne_gau, tr_coord_df = tr_from_to_df_tsne_gau, distance_col = "distance")
trimesh_removed_tsne_gau <- trimesh_removed_tsne_gau +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

df_b <- df_bin_tsne_gau |>
  dplyr::filter(hb_id %in% df_bin_centroids_tsne_gau$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean
## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_tsne_gau$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)
df_exe <- dplyr::bind_rows(df_b, df)
## Set the maximum difference as the criteria
distance_df_small_edges <- distance_tsne_gau |>
  dplyr::filter(distance < benchmark_tsne_gau)
## Since erase brushing is considerd.
langevitour::langevitour(df_exe[1:(length(df_exe)-1)],
                         lineFrom = distance_df_small_edges$from,
                         lineTo = distance_df_small_edges$to,
                         group = df_exe$type, pointSize = append(rep(0, NROW(df_b)), rep(0.5, NROW(df))),
                         levelColors = c("#6a3d9a", "#33a02c"))
```

What can learn from the model:

- There are five clusters. Four of them are Gaussian and two of them are long 2D clusters. Gaussian clusters are close to each other while others are far away.

- Model preserved local and global structure.

- Problem: Couldn't capture the variance within the clusters.

- There is one edge that is not connected in 2D triangular mesh.
